---
title: "Reproducible Analysis of Politician Claims: An R Demo for R2CASS@ICWSM"
author: "Lorraine Saju" # Updated with your name
date: "`r Sys.Date()`"
output: html_document
---

This R Markdown document demonstrates a reproducible analysis of a subset of the LIAR dataset, focusing on politician statements and their fact-checking truth ratings. This analysis is designed to be run seamlessly via Binder, ensuring reproducibility across different environments.

**Goal:** To load a dataset of politician claims, analyze the distribution of their truth ratings, and visualize the results.

## 1. Setup and Data Loading

First, we'll load the necessary libraries (`tidyverse` for data manipulation and plotting, which includes `dplyr`, `ggplot2`, `readr`, etc.). Then we'll load our dataset.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(readr)
library(forcats) # For fct_lump_n, fct_infreq
library(knitr)
library(viridis)
library(RColorBrewer)

# Define column names as per LIAR dataset README, since our TSV has no header
column_names <- c(
  'ID', 'label', 'statement', 'subjects', 'speaker', 'speaker_job_title',
  'state_info', 'party_affiliation', 'barely_true_counts', 'false_counts',
  'half_true_counts', 'mostly_true_counts', 'pants_on_fire_counts', 'context'
)

# Initialize df outside tryCatch for scope
df <- NULL

# Wrap the data loading in tryCatch for conditional messages
tryCatch({
  df <- read_tsv("../data/dataset.tsv", col_names = FALSE) %>%
    setNames(column_names)

  # If successfully loaded and has rows, print success message
  if (!is.null(df) && nrow(df) > 0) {
    cat("Dataset loaded successfully!\n") 
    cat(paste0("Number of claims: ", nrow(df), "\n"))
  } else {
    # This block handles cases where read_tsv succeeds but returns no data
    cat("Error: Dataset loaded but appears empty or invalid.\n")
  }

}, error = function(e) {
  # This block handles actual errors during read_tsv
  cat("Error loading dataset: **", e$message, "**\n\n") # Bold error message
  cat("Please ensure `../data/dataset.tsv` exists and is correctly formatted.\n")
})
```


### Inspecting the Data

Here are the first 5 rows of the dataset:

```{r print_head}
# Ensure df exists before trying to print head
if (!is.null(df) && nrow(df) > 0) {
  df %>%
    head() %>%
    kable(caption = "First 5 Rows of the LIAR Dataset")
} else {
  cat("Data not available for inspection due to loading error.\n")
}
```
## 2. Analyze Distribution of Truth Ratings

Let's examine how many claims fall into each truth rating category (`label` column).

### Distribution of Truth Ratings

```{r truth_rating_distribution}
truth_rating_counts <- df %>%
  count(label) %>%
  arrange(label) %>%
  # Optional: Rename columns for better readability in the table
  rename(`Truth Rating` = label, `Number of Claims` = n)

truth_rating_counts %>%
  kable(caption = "Count of Claims by Truth Rating")
```

## 3. Visualize Truth Rating Distribution
A bar chart provides a clear visual summary of the truth rating distribution. We'll order the labels for better readability.

```{r}
# Define the order for labels to ensure consistent plotting
label_order <- c("pants-fire", "false", "barely-true", "half-true", "mostly-true", "true")

# Create a factor with the desired order
df$label_ordered <- factor(df$label, levels = label_order)

plot_truth_ratings <- df %>%
  ggplot(aes(x = label_ordered, fill = label_ordered)) +
  geom_bar() +
  labs(
    title = "Distribution of Truth Ratings for Politician Claims",
    x = "Truth Rating",
    y = "Number of Claims",
    fill = "Truth Rating"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "RdYlGn")

print(plot_truth_ratings)

# Save the plot
ggsave("truth_ratings_distribution.png", plot_truth_ratings, path = ".", dpi = 300)
```

## 4. Analyze Distribution of Party Affiliation

Next, let's look at the distribution of `party_affiliation` among the speakers.

### Distribution of Party Affiliations

```{r party_affiliation_distribution}
party_affiliation_counts <- df %>%
  count(party_affiliation) %>%
  arrange(desc(n)) %>%
  # Optional: Rename columns for better readability in the table
  rename(`Party Affiliation` = party_affiliation, `Number of Claims` = n)

party_affiliation_counts %>%
  kable(caption = "Count of Claims by Party Affiliation")
```

## 5. Visualize Party Affiliation Distribution
A bar chart to visualize the distribution of speaker party affiliations.

```{r party_affiliation_plot, fig.width=12, fig.height=7}
# Ensure df exists and has data before proceeding with plotting
if (!is.null(df) && nrow(df) > 0) {
  plot_party_affiliation <- df %>%
    # Filter out NA and "none" for cleaner visualization
    filter(!is.na(party_affiliation), party_affiliation != "none") %>%
    mutate(
      # Group parties with small counts into 'Other' category
      # 'n = 10' will keep the 10 most frequent parties and lump the rest
      party_affiliation_grouped = fct_lump_n(fct_infreq(party_affiliation), n = 10, other_level = "Other (Lumped)")
    ) %>%
    ggplot(aes(x = party_affiliation_grouped, fill = party_affiliation_grouped)) + # Use the new grouped factor for x and fill
    geom_bar() +
    labs(
      title = "Distribution of Speaker Party Affiliations (Top 10 + Other)", # Updated title
      x = "Party Affiliation",
      y = "Number of Claims",
      fill = "Party"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_viridis_d(option = "plasma")

  print(plot_party_affiliation)

  # Save the plot
  ggsave("party_affiliation_distribution.png", plot_party_affiliation, path = ".", dpi = 300)
} else {
  cat("Party affiliation plot not generated: Data not available due to loading error.\n")
}
```

## 6. Relationship between Party Affiliation and Truth Ratings
Finally, let's explore how truth ratings vary across different party affiliations using a stacked bar chart.

```{r party_truth_relationship_plot, fig.width=14, fig.height=8}
if (!is.null(df) && nrow(df) > 0) {
  # Apply the same party grouping logic from Section 5 for consistency and readability
  # Also filter out NA and "none" for party affiliations as done previously
  plot_party_vs_truth <- df %>%
    filter(!is.na(party_affiliation), party_affiliation != "none") %>%
    mutate(
      party_affiliation_grouped = fct_lump_n(fct_infreq(party_affiliation), n = 10, other_level = "Other (Lumped)")
    ) %>%
    # Use the ordered truth labels (label_ordered) created in Section 3
    ggplot(aes(x = party_affiliation_grouped, fill = label_ordered)) +
    geom_bar(position = "fill") + # Use position="fill" for proportions (each bar sums to 100%)
    labs(
      title = "Truth Ratings Distribution by Party Affiliation (Proportion, Top 10 Parties)", # Updated title
      x = "Party Affiliation",
      y = "Proportion of Claims",
      fill = "Truth Rating"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right", # Move legend outside the plot area to the right
      legend.box.just = "top" # Align legend to the top of its box
    ) +
    scale_fill_brewer(palette = "RdYlGn") # Red for false, Green for true (consistent with Section 3)

  print(plot_party_vs_truth)

  # Save the plot
  ggsave("party_vs_truth_stacked.png", plot_party_vs_truth, path = ".", dpi = 300)
} else {
  cat("Party affiliation vs. truth rating plot not generated: Data not available due to loading error.\n")
}
```

## Conclusion

This notebook demonstrates how to load, analyze, and visualize data related to politician claims and their truth ratings in a reproducible Python environment. By packaging this with Binder, we ensure anyone can rerun this analysis with ease.

Feel free to modify the code, add new analyses, or explore other aspects of the dataset!

## Session Information

This section provides details about the R session, including the R version and all loaded package versions, to aid in reproducibility.

```{r session_info, echo=TRUE, message=FALSE, warning=FALSE}
sessionInfo()
```
